COVID-19 infection and attributable mortality in UK care homes: Cohort study using active surveillance and electronic records (March-June 2020)

### 2021-02-09 manuscript/manuscript.Rmd
final revisions

residents_episodes <- tline %>% 
  filter(rday==1)  %>% 
  left_join(home_outbreak_sensitivity) %>%
  mutate(
  outbreak = dplyr::case_when(
    date >= first_case_date ~ 1L,
    date < first_case_date ~ 0L,
    TRUE ~ 0L
  )
  # outbreak = dplyr::case_when(
  #   date >= two_case ~ 1L,
  #   date < two_case ~ 0L,
  #   TRUE ~ 0L
  # )
  # outbreak = dplyr::case_when(
  #   date >= five_case ~ 1L,
  #   date < five_case ~ 0L,
  #   TRUE ~ 0L)
  ) %>% 
  group_by(resident_id, home_code, dob, gender, 
           status, home_type, user_type2,
           susceptible_symptomatic, 
           susceptible_confirmed,
           outbreak) %>% 
  summarise(entry_date = min(date),
            exit_date = max(date+1),
            resident_days = sum(rday),
            susceptible_symptomatic_days = sum(susceptible_symptomatic),
            susceptible_confirmed_days = sum(susceptible_confirmed)) %>% 
  mutate(date1 = as.numeric( entry_date - date_min),
         date2 = as.numeric(exit_date - date_min),
         ageg = cut_age_denary(compute_age(dob, date_min), age.min = 75, age.max = 95)) %>% 
  mutate(symptomatic = -susceptible_symptomatic+1,
         confirmed = -susceptible_confirmed+1,
         symptomatic_confirmed = (-susceptible_symptomatic+1)*(-susceptible_confirmed+1),
         date2 = if_else(date1 == date2, date2 + 1, date2)) %>% 
  mutate(covid_status = case_when(
    symptomatic==0 & confirmed == 0 ~ "Uninfected",
    symptomatic==1 & confirmed == 0 ~ "Symptomatic (not confirmed)",
    symptomatic==0 & confirmed == 1 ~ "Asymptomatic confirmed",
    symptomatic==1 & confirmed == 1 ~ "Symptomatic confirmed"
  )) %>% 
  mutate(
    covid_status = factor(
      covid_status, 
      levels = c("Uninfected", "Symptomatic (not confirmed)",
                 "Asymptomatic confirmed", "Symptomatic confirmed")),
    ageg = cut_age_denary(compute_age(dob, date_min), 
                          age.min = 75, age.max = 95))   %>% 
  left_join(
    select(reference_homes, home_code, country, region_UK,
           staff_to_bed_ratio, staff_to_bed_ratio_g5, 
           carer_to_bed_ratio, bed_tot_category, 
           beds, pct_occupancy, no_bedrooms, uk_imd_england_quintile,
           occ_bedroom_ratio, occ_bedroom_ratio_g5, occ_bedroom_ratio_g6)) %>% 
  group_by(resident_id) %>% 
  mutate(death_status = if_else(
    date1 == max(date1),
    as.integer(status == "Deceased"),
    0L
  )) %>% ungroup()

@@

## OUTBREAK HOMES 
desc_residents_outbreak_only <- list()
desc_residents_outbreak_only$outbreak_homes <- unique(
  filter(home_outbreak_sensitivity, !is.na(first_case_date) == 1)$home_code
)

@@

**Results**: `r formatbm(desc_residents$CT_total_symptomatic)`/`r formatbm(desc_residents$CT_total_residents)` residents developed COVID-19 symptoms (`r tbrounding(desc_residents$CT_p_symptomatic)`% [95% confidence interval: `r tbrounding(desc_residents$CT_p_symptomatic_lower)`%; `r tbrounding(desc_residents$CT_p_symptomatic_upper)`%]), while `r formatbm(desc_residents$CT_total_confirmed)` residents (`r tbrounding(desc_residents$CT_p_confirmed)`% [`r tbrounding(desc_residents$CT_p_confirmed_lower)`%; `r tbrounding(desc_residents$CT_p_confirmed_upper)`%]) and `r formatbm(desc_staff$CT_confirmed_total)` staff (`r tbrounding(desc_staff$CT_confirmed_p)`% [`r tbrounding(desc_staff$CT_confirmed_p_lower)`%; `r tbrounding(desc_staff$CT_confirmed_p_upper)`%]) had laboratory-confirmed infections. The rate of incidence of confirmed infection was `r tbrounding(desc_residents$CT_IR_confirmed)` [`r tbrounding(desc_residents$CT_IR_confirmed_lower)`; `r tbrounding(desc_residents$CT_IR_confirmed_upper)`] and `r tbrounding(desc_staff$CT_confirmed_IR)` [`r tbrounding(desc_staff$CT_confirmed_IR_lower)`; `r tbrounding(desc_staff$CT_confirmed_IR_upper)`] per 100,000 person-days in residents and staff respectively. `r length(desc_residents_outbreak_only$outbreak_homes)`/`r desc_residents$total_homes` (`r tbrounding(length(desc_residents_outbreak_only$outbreak_homes)/desc_residents$total_homes*100)`%) care homes had at least one COVID-19 infection or COVID-19-related death. Lower staffing ratios and higher occupancy rates were independent risk factors for infection.

`r desc_residents$DAT_CFR$y` of `r desc_residents$DAT_CFR$n` residents with confirmed infection died (case-fatality rate: `r tbrounding(desc_residents$DAT_CFR_rate)`% [`r tbrounding(desc_residents$DAT_CFR_rate_lower)`%; `r tbrounding(desc_residents$DAT_CFR_rate_upper)`%]). Mortality in residents with no direct evidence of infection was two-fold higher in care homes with outbreaks versus those without (adjusted HR 2.1 [1.7; 2.5]).

@@

Globally the number of COVID-19 cases continues to increase, with substantially higher rates of infection reported in both care homes and hospitals.[@PHE2020] In the UK, there are an estimated 400,000 residents living in approximately 11,000 care homes for older people with or within on-site nursing care.[@LaingBuisson2019; @Sanford2015] Care home residents are particularly vulnerable to COVID-19 due to older age, high prevalence of comorbidity,[@Zhou2020] and frequent exposure to infection through contact with staff, other residents and contaminated surfaces in the facility. At the peak of the pandemic, deaths recorded in UK care home residents were three times higher than during the preceding year.[@onsDR2020] Staff in UK care homes also had higher aged-standardised rates of COVID-19 related mortality compared to other occupations.[@onsDR2020occ] UK national statistics suggests two-thirds of excess deaths recorded in care home residents in the last 6 months involved COVID-19,[@onsDR2020] but this is likely to be an underestimate because many residents were not tested. Understanding the proportion of excess deaths that can be directly and indirectly attributed to COVID-19 infection is important, to fully assess the impact of the pandemic on care homes.

@@

To our knowledge, there are no studies which have employed population-level active surveillance (daily monitoring to identify possible cases of COVID-19 in residents and staff) in care homes to investigate the epidemiology and clinical outcomes of both suspected and confirmed COVID-19 infections. We analysed electronic health records from the Four Seasons Health Care Group, one of the UK’s largest for-profit provider of residential and nursing care, with the aim of identifying strategies to protect staff and residents in care homes from future waves of infection. Our objectives were to estimate incidence of and risk factors for infection, and incidence of mortality in the following groups: (A) residents with no evidence of infection; (B) symptomatic residents; (C) asymptomatic residents with confirmed infection; and (D) symptomatic residents with confirmed infection. We also estimated mortality attributable to COVID-19.

@@

To adjust for case under-ascertainment, we created a dummy variable indicating 'affected care homes', which we define throughout the manuscript as care homes that recorded $\geq$1 confirmed infection or COVID-19 related death during the period.

@@

In order to investigate the relationship between COVID-19 infection and excess mortality, we made the assumption that residents in 'non-affected' care homes (no record of any confirmed cases or COVID-19 related deaths) had not been exposed to infection, and would therefore not experience excess COVID-19 related mortality.[@Burton2020] We therefore compared all-cause mortality in residents with no evidence of infection (group A) in care homes with and without outbreaks.

@@

Time-dependent Cox proportional hazard models (`r table_nums("tab_CoxPH_death", display = "cite")`) examine the relationship between infection status (groups A–D) and mortality mortality (Figure S3). After controlling for other risk factors, increased mortality was independently associated with older age, male gender (adjusted HR = 1.5 [1.3; 1.6]), and receiving nursing care (adjusted HR = 1.4 [1.1; 1.6]).

We estimated excess mortality in outbreak and non-outbreak care homes, taking individuals with no evidence of infection (group A) in non-outbreak care homes as the reference group. Hazards of all-cause mortality were two-fold higher in Group A – no direct evidence of infection in outbreak versus non-outbreak care homes (adjusted HR = 2.1 [1.7; 2.5]). All-cause mortality was strongly associated with confirmed infection, whether asymptomatic (Group C: adjusted HR = 3.4 [2.1; 5.7]) or symptomatic (Group D: adjusted HR = 13 [10; 16]). In confirmed infections, mortality was significantly higher in individuals with a record of symptoms.

@@

```{r hazard_ratios_death,  warning=F, results="asis"}
# foreign::write.dta(residents_episodes, 
#                    file = "../residents_episodes.dta")
# cph_death <- coxph(
#   Surv(date1, date2, death_status) ~ gender + ageg +
#     home_type + user_type2 + bed_tot_category +
#     uk_imd_england_quintile +
#     I(bed_to_staff_ratio-.85) + I(occ_bedroom_ratio-.9) +
#     covid_status_outbreak_digit1 +
#     covid_status_outbreak_digit2 +
#     covid_status_outbreak_digit3 +
#     covid_status_outbreak_digit4 +
#     covid_status_outbreak_digit5 +
#     + cluster(resident_id),
#   data = residents_episodes)
# cox.zph(cph_death)
# for(i in 8:12){
#   plot(cox.zph(cph_death)[i], col = 2)
# }
explanatory = c(
  "gender", "ageg", "home_type", "user_type2", 
  "bed_tot_category", "uk_imd_england_quintile",
  "I(occ_bedroom_ratio-.9)", "I(bed_to_staff_ratio-.85)",
  "covid_status_outbreak",  "cluster(home_code)")
dependent = "Surv(time = date1, time2 = date2+0.1, event = death_status)"
CPH_death <-  finalfit(residents_episodes, dependent, 
                       explanatory, column = T, 
                       add_dependent_label = F)  %>% 
  ff_remove_p() 
knitr::kable(CPH_death, row.names=FALSE, 
             align=c("l", "l", "r", "r", "r", "r"))
```

@@

Model-based estimates indicate that 633/1,694 (37%) all-cause deaths were attributable to COVID-19 (Table S7). In care homes with outbreaks only, just 161/609 (26%) deaths attributa-ble to COVID-19 occurred in people with confirmed infection (Groups C and D, Table S7).

@@

```{r attributable_deaths, warning=F, message=F, results = "asis"}
cph_death3 <- coxph(Surv(date1, date2+0.1, death_status) ~
                    gender+ ageg + home_type + user_type2 + covid_status_outbreak_digit +
                    bed_tot_category + uk_imd_england_quintile +
                    I(occ_bedroom_ratio-.9) + I(bed_to_staff_ratio-.85) +
                    cluster(home_code), data =  residents_episodes)

@@

attributable_deaths <- residents_episodes %>%
  group_by(covid_status_outbreak_digit, covid_status_outbreak) %>%
  summarise(
    rdays = sum(as.integer(date2-date1)),
    deaths = sum(death_status == 1),
    deaths_apr = sum(death_status == 1 & date2 >= 30)
    ) %>% 
  left_join(HRs_covid) %>% 
  mutate(HR = tidyr::replace_na(HR, 1),
           HR_lower = tidyr::replace_na(HR_lower, 1),
           HR_upper = tidyr::replace_na(HR_upper, 1))

@@

attributable_deaths <- bind_rows(attributable_deaths, 
                                 colSums(attributable_deaths[,c("rdays", "deaths", "deaths_apr",
                                                                "AF", "AF_apr")]))
attributable_deaths %>% ungroup() %>% transmute(
  `Infection/outbreak status` = tidyr::replace_na(as.character(covid_status_outbreak), "TOTAL"),
  `Adjusted HR` = paste0(tbrounding(HR,2), " [",
                        tbrounding(HR_lower,2),"; ",
                        tbrounding(HR_upper,2), "]"),
  `Resident-days` = formatbm(rdays),
  `Total deaths` = formatbm(deaths),
  # `Total deaths (April-June)` = formatbm(deaths_apr),
  `Deaths attributable to COVID-19 (% all-cause deaths)` = paste0(
    round(AF * sum(deaths/2)), " (",
    tbrounding(AF*100, 2), ")"),
  # `Deaths attributable to COVID-19 (% all-cause deaths April-June)` = paste0(
  #   round(AF_apr * sum(deaths_apr/2)), " (",
  #   tbrounding(AF_apr*100, 2), ")")
) %>% kable(align=c("l", rep("r", 4)))

@@

Across 179 care homes, 22% of residents experienced symptoms while 10% had a laboratory-confirmed infection, with a case-fatality rate of 35.7% across the first wave of the pandemic. Residents with no direct evidence of infection in care homes with outbreaks had twice the mortality of the equivalent group in care homes without outbreaks: less than one-third of deaths attributable to COVID-19 in outbreak care homes were confirmed due to insufficient testing capacity until late in the pandemic. In addition to the need for active surveillance linked to increased testing capacity, higher staff to resident ratios and reduced care home occupancy may be important to reducing the spread of infection.

@@

In common with a Canadian cohort study,[@Brown2020] we found strong associations between infections and care home baseline occupancy. We also found staffing levels to be negatively associated with infection rates. These organisational factors, linked to chronic underfunding of the care sector, are likely to facilitate the implementation of infection control procedures[@dhICguideline2020] such as isolating or cohorting infected residents, staff training, and regular environmental deep cleaning. When staff care for fewer residents they also have reduced likelihood of spreading infection between residents. Higher staff to resident ratios may also decrease reliance on agency staff who often work across multiple settings, and indicate better resourced care homes. This association may also be confounded by other characteristics of homes with higher staffing levels, if those happen to be better resourced, for instance in relation to staff training, availability of personal protective equipment, building characteristics.

@@

Internal validity of our findings on mortality attributable to COVID-19 is sensitive to the assumption made that residents of care homes categorised as 'not affected' were not exposed to infections. Other key limitations relate to the completeness of individual-level data recorded through Datix, which we used for all individual-level analyses. We lacked information on comorbidity and ethnicity, both of which have been shown to be important risk factors for adverse outcomes in COVID-19,[@Zhou2020] but we were able to identify individuals with dementia, and adjust for receipt of nursing care which will partially capture comorbidity. The number of infections was under-reported in the individual-level dataset by comparison with the manager-reported daily infection tallies, and we lacked information on the overall rate of testing in each care home.

>>>>

residents_episodes <- tline %>% 
  filter(rday==1)  %>% 
  left_join(home_outbreak_sensitivity) %>%
  mutate(
  outbreak = dplyr::case_when(
    date >= first_case_date ~ 1L,
    date < first_case_date ~ 0L,
    TRUE ~ 0L
  )
  # outbreak = dplyr::case_when(
  #   date >= two_case ~ 1L,
  #   date < two_case ~ 0L,
  #   TRUE ~ 0L
  # )
  # outbreak = dplyr::case_when(
  #   date >= five_case ~ 1L,
  #   date < five_case ~ 0L,
  #   TRUE ~ 0L)
  ) %>% 
  group_by(resident_id, home_code, dob, gender, 
           status, home_type, user_type2,
           susceptible_symptomatic, 
           susceptible_confirmed,
           outbreak) %>% 
  summarise(entry_date = min(date),
            exit_date = max(date)+.99,
            resident_days = sum(rday),
            susceptible_symptomatic_days = sum(susceptible_symptomatic),
            susceptible_confirmed_days = sum(susceptible_confirmed)) %>% 
  mutate(date1 = as.numeric(entry_date - date_min),
         date2 = as.numeric(exit_date - date_min),
         ageg = cut_age_denary(compute_age(dob, date_min), age.min = 75, age.max = 95)) %>% 
  mutate(symptomatic = -susceptible_symptomatic+1,
         confirmed = -susceptible_confirmed+1,
         symptomatic_confirmed = (-susceptible_symptomatic+1)*(-susceptible_confirmed+1),
         date2 = if_else(date1 == date2, date2 + .2, date2)) %>% 
  mutate(covid_status = case_when(
    symptomatic==0 & confirmed == 0 ~ "Uninfected",
    symptomatic==1 & confirmed == 0 ~ "Symptomatic (not confirmed)",
    symptomatic==0 & confirmed == 1 ~ "Asymptomatic confirmed",
    symptomatic==1 & confirmed == 1 ~ "Symptomatic confirmed"
  )) %>% 
  mutate(
    covid_status = factor(
      covid_status, 
      levels = c("Uninfected", "Symptomatic (not confirmed)",
                 "Asymptomatic confirmed", "Symptomatic confirmed")),
    ageg = cut_age_denary(compute_age(dob, date_min), 
                          age.min = 75, age.max = 95))   %>% 
  left_join(
    select(reference_homes, home_code, country, region_UK,
           staff_to_bed_ratio, staff_to_bed_ratio_g5, 
           carer_to_bed_ratio, bed_tot_category, 
           beds, pct_occupancy, no_bedrooms, uk_imd_england_quintile,
           occ_bedroom_ratio, occ_bedroom_ratio_g5, occ_bedroom_ratio_g6)) %>% 
  group_by(resident_id) %>% 
  mutate(death_status = if_else(
    date1 == max(date1),
    as.integer(status == "Deceased"),
    0L
  )) %>% ungroup()

@@

## OUTBREAK HOMES 
desc_residents_outbreak_only <- list()
desc_residents_outbreak_only$outbreak_homes <- unique(
  filter(home_outbreak_sensitivity, !is.na(
    # two_case
    first_case_date
    ) == 1)$home_code
)

@@

**Results**: `r formatbm(desc_residents$CT_total_symptomatic)`/`r formatbm(desc_residents$CT_total_residents)` residents developed COVID-19 symptoms (`r tbrounding(desc_residents$CT_p_symptomatic)`% [95% confidence interval: `r tbrounding(desc_residents$CT_p_symptomatic_lower)`%; `r tbrounding(desc_residents$CT_p_symptomatic_upper)`%]), while `r formatbm(desc_residents$CT_total_confirmed)` residents (`r tbrounding(desc_residents$CT_p_confirmed)`% [`r tbrounding(desc_residents$CT_p_confirmed_lower)`%; `r tbrounding(desc_residents$CT_p_confirmed_upper)`%]) and `r formatbm(desc_staff$CT_confirmed_total)` staff (`r tbrounding(desc_staff$CT_confirmed_p)`% [`r tbrounding(desc_staff$CT_confirmed_p_lower)`%; `r tbrounding(desc_staff$CT_confirmed_p_upper)`%]) had laboratory-confirmed infections. The incidence of confirmed infection was `r tbrounding(desc_residents$CT_IR_confirmed)` [`r tbrounding(desc_residents$CT_IR_confirmed_lower)`; `r tbrounding(desc_residents$CT_IR_confirmed_upper)`] and `r tbrounding(desc_staff$CT_confirmed_IR)` [`r tbrounding(desc_staff$CT_confirmed_IR_lower)`; `r tbrounding(desc_staff$CT_confirmed_IR_upper)`] per 100,000 person-days in residents and staff respectively. `r length(desc_residents_outbreak_only$outbreak_homes)`/`r desc_residents$total_homes` (`r tbrounding(length(desc_residents_outbreak_only$outbreak_homes)/desc_residents$total_homes*100)`%) care homes had at least one COVID-19 infection or COVID-19-related death. Lower staffing ratios and higher occupancy rates were independent risk factors for infection.

`r desc_residents$DAT_CFR$y` of `r desc_residents$DAT_CFR$n` residents with confirmed infection died (case-fatality rate: `r tbrounding(desc_residents$DAT_CFR_rate)`% [`r tbrounding(desc_residents$DAT_CFR_rate_lower)`%; `r tbrounding(desc_residents$DAT_CFR_rate_upper)`%]). Mortality in residents with no direct evidence of infection was two-fold higher in care homes with outbreaks versus those without (adjusted HR 2.2 [1.8; 2.6]).

@@

Globally the number of COVID-19 cases continues to increase, with substantially higher rates of infection reported in both care homes and hospitals.[@PHE2020] In the UK, there are an estimated 400,000 residents living in approximately 11,000 care homes for older people, which provide residential care with or without on-site nursing.[@LaingBuisson2019; @Sanford2015] Care home residents are particularly vulnerable to COVID-19 due to older age, high prevalence of comorbidity,[@Zhou2020] and frequent exposure to infection through contact with staff, other residents and contaminated surfaces in the facility. At the peak of the pandemic, deaths recorded in UK care home residents were three times higher than during the preceding year.[@onsDR2020] Staff in UK care homes also had higher aged-standardised rates of COVID-19 related mortality compared to other occupations.[@onsDR2020occ] UK national statistics suggests two-thirds of excess deaths recorded in care home residents in the last 6 months involved COVID-19,[@onsDR2020] but this is likely to be an underestimate because many residents were not tested. Understanding the proportion of excess deaths that can be directly and indirectly attributed to COVID-19 infection is important, to fully assess the impact of the pandemic on care homes.

@@

To our knowledge, there are no studies which have employed population-level active surveillance (daily monitoring to identify possible cases of COVID-19 in residents and staff) in care homes to investigate the epidemiology and clinical outcomes of both suspected and confirmed COVID-19 infections. We analysed electronic health records from the Four Seasons Health Care Group, one of the UK’s largest for-profit providers of residential and nursing care, with the aim of identifying strategies to protect staff and residents in care homes from future waves of infection. Our objectives were to estimate incidence of and risk factors for infection, and incidence of mortality in the following groups: (A) residents with no evidence of infection; (B) symptomatic residents; (C) asymptomatic residents with confirmed infection; and (D) symptomatic residents with confirmed infection. We also estimated mortality attributable to COVID-19.

@@

To adjust for case under-ascertainment, we created a dummy variable indicating care homes that had experienced an outbreak, defined throughout the manuscript as care homes that recorded ≥1 confirmed infection or COVID-19 related death during the study period. This definition was preferred over a standard outbreak definition (≥2 cases linked in time and place) to compensate for poor COVID-19 case-ascertainment during the pandemic due to limited testing. Sensitivity analysis was used to assess the impact on excess mortality of using a more specific outbreak definition (supplementary methods).

@@

In order to investigate the relationship between COVID-19 infection and excess mortality, we made the assumption that residents in 'non-outbreak' care homes (no record of any confirmed cases or COVID-19 related deaths) had not been exposed to infection, and would therefore not experience excess COVID-19 related mortality.[@Burton2020] We therefore compared all-cause mortality in residents with no evidence of infection (group A) in care homes with and without outbreaks.

@@

Time-dependent Cox proportional hazard models (`r table_nums("tab_CoxPH_death", display = "cite")`) examine the relationship between infection status (groups A–D) and mortality (Figure S3). After controlling for other risk factors, increased mortality was independently associated with older age, male gender (adjusted HR = 1.5 [1.3; 1.6]), and receiving nursing care (adjusted HR = 1.3 [1.1; 1.6]).

We estimated excess mortality in outbreak and non-outbreak care homes, taking individuals with no evidence of infection (group A) in non-outbreak care homes as the reference group. Hazards of all-cause mortality were two-fold higher in Group A – no direct evidence of infection in outbreak versus non-outbreak care homes (adjusted HR = 2.2 [1.8; 2.6]). All-cause mortality was strongly associated with confirmed infection, whether asymptomatic (Group C: adjusted HR = 3.8 [2.3; 6.4]) or symptomatic (Group D: adjusted HR = 14 [11; 18]). In confirmed infections, mortality was significantly higher in individuals with a record of symptoms.

@@

```{r hazard_ratios_death,  warning=F, results="asis"}
# foreign::write.dta(residents_episodes, 
#                    file = "../residents_episodes.dta")
# cph_death <- coxph(
#   Surv(date1, date2, death_status) ~ gender + ageg +
#     home_type + user_type2 + bed_tot_category +
#     uk_imd_england_quintile +
#     I(bed_to_staff_ratio-.85) + I(occ_bedroom_ratio-.9) +
#     covid_status_outbreak_digit1 +
#     covid_status_outbreak_digit2 +
#     covid_status_outbreak_digit3 +
#     covid_status_outbreak_digit4 +
#     covid_status_outbreak_digit5 +
#     + cluster(resident_id),
#   data = residents_episodes)
# cox.zph(cph_death)
# for(i in 8:12){
#   plot(cox.zph(cph_death)[i], col = 2)
# }
explanatory = c(
  "gender", "ageg", "home_type", "user_type2", 
  "bed_tot_category", "uk_imd_england_quintile",
  "I(occ_bedroom_ratio-.9)", "I(bed_to_staff_ratio-.85)",
  "covid_status_outbreak",  "cluster(home_code)")
dependent = "Surv(time = date1, time2 = date2, event = death_status)"

CPH_death <-  finalfit(residents_episodes, dependent, 
                       explanatory, column = T, 
                       add_dependent_label = F)  %>% 
  ff_remove_p() 
knitr::kable(CPH_death, row.names=FALSE, 
             align=c("l", "l", "r", "r", "r", "r"))
```

@@

Model-based estimates indicate that 653/1,694 (39%) all-cause deaths were attributable to COVID-19 (Table S7). In care homes with outbreaks only, just 217/653 (33%) deaths attributable to COVID-19 occurred in people with confirmed infection (Groups C and D, Table S7).

@@

```{r attributable_deaths, warning=F, message=F, results = "asis"}
cph_death3 <- coxph(Surv(date1, date2, death_status) ~
                    gender+ ageg + home_type + user_type2 + covid_status_outbreak_digit +
                    bed_tot_category + uk_imd_england_quintile +
                    I(occ_bedroom_ratio-.9) + I(bed_to_staff_ratio-.85) +
                    cluster(home_code), data =  residents_episodes)

@@

attributable_deaths <- residents_episodes %>%
  group_by(covid_status_outbreak_digit, covid_status_outbreak) %>%
  summarise(
    rdays = sum(date2-date1),
    deaths = sum(death_status == 1),
    deaths_apr = sum(death_status == 1 & date2 >= 30)
    ) %>% 
  left_join(HRs_covid) %>% 
  mutate(HR = tidyr::replace_na(HR, 1),
           HR_lower = tidyr::replace_na(HR_lower, 1),
           HR_upper = tidyr::replace_na(HR_upper, 1))

@@

attributable_deaths <- bind_rows(attributable_deaths, 
                                 colSums(attributable_deaths[,c("rdays", "deaths", "deaths_apr",
                                                                "AF", "AF_apr")]))
attributable_deaths %>% ungroup() %>% transmute(
  `Infection/outbreak status` = tidyr::replace_na(as.character(covid_status_outbreak), "TOTAL"),
  `Adjusted HR` = paste0(tbrounding(HR,2), " [",
                        tbrounding(HR_lower,2),"; ",
                        tbrounding(HR_upper,2), "]"),
  `Resident-days` = formatbm(rdays),
  `Deaths attributable to COVID-19 (% all-cause deaths)` = round(AF * sum(deaths/2)),
  `Total deaths` = formatbm(deaths),
  `%` = tbrounding(AF*100, 2)

  # `Total deaths (April-June)` = formatbm(deaths_apr),
  # `Deaths attributable to COVID-19 (% all-cause deaths April-June)` = paste0(
  #   round(AF_apr * sum(deaths_apr/2)), " (",
  #   tbrounding(AF_apr*100, 2), ")")
) %>% kable(align=c("l", rep("r", 4)))

@@

Across 179 care homes, 22% of residents experienced symptoms while 10% had a laboratory-confirmed infection, with a case-fatality rate of 35.7% across the first wave of the pandemic. Residents with no direct evidence of infection in care homes with outbreaks had twice the mortality of the equivalent group in care homes without outbreaks: less than one-third of deaths attributable to COVID-19 in outbreak care homes were confirmed due to insufficient testing capacity until late in the pandemic. In addition to the need for active surveillance linked to increased testing capacity, higher staff to resident ratios and reduced care home occupancy may be important to reduce the spread of infection.

@@

In common with a Canadian cohort study,[@Brown2020] we found strong associations between infections and care home baseline occupancy. We also found staffing levels to be negatively associated with infection rates. These organisational factors, linked to chronic underfunding of the care sector, are likely to hinder the implementation of robust infection control procedures[@dhICguideline2020] such as isolating or cohorting infected residents, staff training, and regular environmental deep cleaning. When staff care for fewer residents they also have reduced likelihood of spreading infection between residents. Higher staff to resident ratios may also decrease reliance on agency staff who often work across multiple settings, and indicate better resourced care homes. These associations may also be confounded by other care home characteristics which could not be measured in this study, such as access to personal protective equipment, or the structure and layout of the building.

@@

Our estimates of mortality attributable to COVID-19 are dependent on our definition of 'outbreak' versus 'non-outbreak' care homes. We used a sensitive definition of outbreaks (≥1 case per care home) because case under-ascertainment in care homes was common due to lack of testing.  However, it is possible that we incorrectly classified some care homes with one or a handful of cases throughout the pandemic as having experienced outbreaks. Other key limitations relate to the completeness of individual-level data recorded through Datix, which we used for all individual-level analyses. We lacked information on comorbidity and ethnicity, both of which have been shown to be important risk factors for adverse outcomes in COVID-19,[@Zhou2020] but we were able to identify individuals with dementia, and adjust for receipt of nursing care which will partially capture comorbidity. The number of infections was under-reported in the individual-level dataset by comparison with the manager-reported daily infection tallies, and we lacked information on the overall rate of testing in each care home. In addition, our measures of care home occupancy were based on the pre-pandemic period and did not take account of higher vacancy rates during follow-up.
